---
title: "Track 01 Starter"
output: html_document
---

<style>
body { font-family: 'DM Sans', 'Helvetica Neue', sans-serif; }
</style>

# Track 01 Starter

Scaffold for importing the Track 01 data package and checking integrity before plotting. No contrast is selected and no final result is computed.

## Load the data

```{r}
counts_path <- "../../data/track-01/track01_counts_vt.csv"
metadata_path <- "../../data/track-01/track01_metadata_vt.csv"

read_or_pending <- function(path) {
  if (file.exists(path) && file.info(path)$size > 0) {
    read.csv(path, row.names = 1)
  } else {
    message(basename(path), " is pending HDAG release.")
    NULL
  }
}

counts <- read_or_pending(counts_path)
metadata <- read_or_pending(metadata_path)

if (!is.null(counts)) print(dim(counts))
if (!is.null(metadata)) print(dim(metadata))
```

## Integrity checks

Confirm the data are what they claim to be before any analysis. Do not silently fix values that fail a check. Report them in your interpretation memo.

```{r}
if (!is.null(counts) && !is.null(metadata)) {
  cat("samples in both files:", length(intersect(rownames(metadata), colnames(counts))), "\n")
  cat("counts are nonnegative:", all(as.matrix(counts) >= 0), "\n")
  cat("counts are integral:", all(as.matrix(counts) %% 1 == 0), "\n")
  cat("duplicate genes:", sum(duplicated(rownames(counts))), "\n")
  cat("duplicate samples:", sum(duplicated(colnames(counts))), "\n")
}
```

## DESeq2 skeleton

Fill in the design and contrast from the challenge guide. This chunk is deliberately not evaluated.

```{r, eval=FALSE}
library(DESeq2)

dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(counts),
  colData = metadata,
  design = ~ group
)
dds <- DESeq(dds)
res <- results(dds, contrast = c("group", "case", "control"))
```

## Exploratory plots

```{r, eval=FALSE}
library(ggplot2)

vsd <- vst(dds)
plotPCA(vsd, intgroup = "group")

sample_dists <- dist(t(assay(vsd)))
pheatmap::pheatmap(as.matrix(sample_dists))
```
